A kernel for incremental type-checking

### 2010-05-11 doc/typing_rules.tex
Definition de la deuxième version des règles de typages.
On a simplifié un certain nombre de caractéristiques inutiles du langage
précédent: fusion atome/transformer, plus d'arité mais un système
curryfié, plus de conclusion multiple possible.

git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@41 403e655e-980a-47bd-8c60-1ee5348bb6b7

\documentclass[12pt]{article}
\usepackage{mathpartir}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{mathpartir}
\usepackage{color}
%\usepackage[pdftex,backref=page,colorlinks=true]{hyperref}
\usepackage{amsmath, amstext, amsthm, amsfonts}
\usepackage{stmaryrd}
\usepackage[protrusion=true,expansion=true]{microtype}

\title{Typing rules for the data-oriented patch system}

@@

\begin{align*}
 \gdecl{a}{x \gor a\ x } \\
 \gdecl{t}{a \gor \sort \gor (x:t)\to t \gor (x=a:t)\to t} \\
 \gdecl{\Gamma}{\cdot \gor \Gamma[x:t] \gor \Gamma[x=a:t]}
\end{align*}

Following Agda's syntax, we denote by $(x:t)\to u$ the dependent
product (elsewhere written with a $\Pi$). The new construct
$(x=a:t)\to u$ is a singleton type in argument position, whose value
is $a$. Such a singleton product should be read as ``the function
taking as argument any values syntactically equal to $a$ of type $t$
(\emph{i.e.} a pedantic way to say $a$). We write as usual
$(x_1:t_1)(x_2:t_2)\ldots(x_n:t_n)\to t$ for
$(x_1:t_1)\to(x_2:t_2)\to\ldots\to(x_n:t_n)\to t$. The lookup in the
environment is written either $\Gamma(x):t$, denoting ``$\Gamma$
contains a declaration $[x:t]$ \emph{or} a definition $[x=a:t]$'', or
$\Gamma(x)=a:t$ denoting ``$\Gamma$ contains a definition
$[x=a:t]$''. Capture-avoiding renaming of variables in a term $t$ is
written $t\subst{x}{y}$ for ``replace all occurences of $x$ by $y$''
and is defined the usual way.

We maintain the strong invariant that the environment does not contain
two variables of the same name, nor two syntactically equal
declarations (the $a$ part is unique). More on that in the
implementation part.

\paragraph{Typing} The typing of our system is a restriction of the
PTS's typing rules with the additional two rules for typing a singleton
product and the application of a singleton argument (featuring the
syntactic verification in the environment). It relies on two distincts
jugements:
\begin{itemize}
\item $\Gamma\vdash a : t$ ($a$ has type $t$ in $\Gamma$),
\item $\Gamma\vdash t : \sort$ ($t$ has sort $s$ in $\Gamma$).
\end{itemize}
The rules are:
\begin{mathpar}
  \infer[Axiom]{
  }{\Gamma\vdash\sort_1:\sort_2}
  \quad(\sort_1,\sort_2)\in\mathcal A
  \and
  \infer[Init]{
    \Gamma(x):t
  }{\Gamma\vdash x:t}
  \and
  \infer[Prod]{\Gamma\vdash t:\sort_1 \and 
    \Gamma[x:t]\vdash u:\sort_2
  }{\Gamma\vdash(x:t)\to u : \sort_3}
  \quad(\sort_1,\sort_2,\sort_3)\in\mathcal R
  \and
  \infer[Sing]{\Gamma\vdash a:t \and
    \Gamma\vdash t:\sort_1 \and
    \Gamma[x=a:t]\vdash u:\sort_2
  }{\Gamma\vdash(x=a:t)\to u : \sort_3} 
  \quad(\sort_1,\sort_2,\sort_3)\in\mathcal R
  \and
  \infer[AppP]{
    \Gamma\vdash a:(y:t)\to u \and \Gamma(x):t
    }{\Gamma\vdash a\ x : u\subst{x}{y}}
  \and
  \infer[AppS]{
    \Gamma\vdash a:(y=b:t)\to u \and \Gamma(x)=b:t
    }{\Gamma\vdash a\ x : u\subst{x}{y}}
\end{mathpar}

@@

The main problem when it comes with implementation is to guarantee the
unicity of declarations and definitions in the environment. One
approach would be to use the locally-nameless method, along with a
special type of free names: A term is stored in deBruijn syntax, but
as soon as a variable is introduced, it takes a real name.

@@

>>>>

\documentclass[12pt]{article}
\usepackage{mathpartir}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{mathpartir}
\usepackage{color}
%\usepackage[pdftex,backref=page,colorlinks=true]{hyperref}
\usepackage{amsmath, amstext, amsthm, amsfonts}
\usepackage{stmaryrd}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{fullpage}

\title{A kernel for incremental type-checking}

@@

\begin{align*}
 \gdecl{a}{x \gor a\ x } \\
 \gdecl{t}{a \gor \sort \gor (x:t)\cdot t \gor (x=a:t)\cdot t} \\
 \gdecl{\Gamma}{\cdot \gor \Gamma[x:t] \gor \Gamma[x=a:t]}
\end{align*}

Following the standard syntax, we denote by $(x:t)\cdot u$ the
dependent product (elsewhere written with a $\Pi$). The new construct
$(x=a:t)\cdot u$ is a kind of singleton type, whose value is $a$. Such
a singleton product should be read as ``the (implicit) function taking
as argument any values syntactically equal to $a$ of type $t$
(\emph{i.e.} a pedantic way to say $a$).

We write as usual $(x_1:t_1)(x_2:t_2)\ldots(x_n:t_n)\cdot t$ for
$(x_1:t_1)\cdot(x_2:t_2)\cdot\ldots\cdot(x_n:t_n)\cdot t$, $t \to
u$ for $(x:t)\cdot u$ where $x\notin u$, and $(x\ y : t)\cdot u$ for
$(x:t)(y:t)\cdot u$. The lookup in the environment
is written either $\Gamma(x):t$, denoting ``$\Gamma$ contains a
declaration $[x:t]$ \emph{or} a definition $[x=a:t]$'', or
$\Gamma(x)=a:t$ denoting ``$\Gamma$ contains a definition
$[x=a:t]$''. Capture-avoiding renaming of variables in a term $t$ is
written $t\subst{x}{y}$ for ``replace all occurences of $x$ by $y$''
and is defined the usual way.

We maintain the strong invariant that the environment does not contain
two variables of the same name, nor two syntactically equal
definitions (the $a$ part is unique). More on that in the
implementation part.

\paragraph{Typing} The typing of our system is a restriction of the
PTS's typing rules with the additional two rules for typing a singleton
product and the application of a singleton argument (featuring the
syntactic verification in the environment). It relies on two distincts
jugements:
\begin{itemize}
\item $\Gamma\vdash a : t$ ($a$ has type $t$ in $\Gamma$),
\item $\Gamma\vdash t : \sort$ ($t$ has sort $s$ in $\Gamma$).
\end{itemize}
The rules are:
\begin{mathpar}
  \infer[Axiom]{
  }{\Gamma\vdash\sort_1:\sort_2}
  \quad(\sort_1,\sort_2)\in\mathcal A
  \and
  \infer[Init]{
    \Gamma(x):t
  }{\Gamma\vdash x:t}
  \and
  \infer[Prod]{\Gamma\vdash t:\sort_1 \and 
    \Gamma[x:t]\vdash u:\sort_2
  }{\Gamma\vdash(x:t)\cdot u : \sort_3}
  \quad(\sort_1,\sort_2,\sort_3)\in\mathcal R
  \and
  \infer[Sing]{\Gamma\vdash a:t \and
    \Gamma\vdash t:\sort_1 \and
    \Gamma[x=a:t]\vdash u:\sort_2
  }{\Gamma\vdash(x=a:t)\cdot u : \sort_3} 
  \quad(\sort_1,\sort_2,\sort_3)\in\mathcal R
  \and
  \infer[App]{
    \Gamma\vdash a:(y:t)\cdot u \and \Gamma(x):t
    }{\Gamma\vdash a\ x : u\subst{x}{y}}
  \and
  \infer[Skip]{
    \Gamma\vdash a:(y=b:t)\cdot u \and \Gamma(x)=b:t
    }{\Gamma\vdash a : u\subst{x}{y}}
\end{mathpar}

@@

The main problem when it comes with implementation is to guarantee the
unicity of declarations and definitions in the environment. One
approach would be to use the locally-nameless method, along with a
special type of free names: A term is stored in deBruijn syntax, but
as soon as a variable is introduced, it takes a real name. So in fact,
environments are not syntactic list objects, instead the operations
$\Gamma[x:t]$ and $\Gamma[x=a:t]$ are the partial meta-operations of
adding new entries into the environment, which returns a new, unique
name for them. The last meta step is to substitute $x$ for this new
name.

@@

Let's put aside the problem of configuring the PTS for the moment,
since it's not clear yet: we work in system U \emph{i.e.}
$\langle\{*\},\{*:*\},\{(*,*,*)\}\rangle$.

We are going to construct a repository gradually as a big term, but
stopping at some steps to explain the construct. The beginning of the
term will consist in the declaration of the language (a small
arithmetic language). Keep in mind that this simply typed language is
just an example: more evolved languages can and will be declared, even
on-the-fly as part of a commit!
\newcommand\nat{\mathbb N}
\newcommand\prop{prop}
\begin{align}&
  (\prop : *)(\nat : *)(0:\nat)(S:\nat) \\&
  (\land\ \lor\Rightarrow\ : \prop \to \prop \to \prop) \\&
  (+\ -\ : \nat\to\nat\to\nat)
  (=\ \neq\ \leq:\nat\to\nat\to \prop )
  \intertext{Now let's begin the definition of the expression
    $2+2=4$. The syntax restricts us to define it step-by-step:} &
  (1 = S\ 0 : \nat)(2 = S\ 1 : \nat)(3 = S\ 2 :
  \nat)(4=S\ 3 : \nat) \\&
  (t = (+)\ 2\ 2 : \nat) (p : (=)\ t\ 4) \\&
  \intertext{Forever, the name $p$ will be associated to that
    expression. Let's mark it with a name. First we introduce -- on
    the fly -- the marking feature to our repository (how cool is that?):} &
  (commit : *)
  (mark : \prop\to string\to commit)
  \intertext{The type $string$ is predefined in some way, and the
    choice of $prop$ as the argument to $mark$ fixes that we want to
    record propositions as commits, not naturals: it is in some sense
    the ``tip'' of our syntax. Now we can mark, or commit that $p$: } &
  (x = mark\ p\ \mathtt{``Cool\ equation}" : commit)
  \intertext{Now let's make it evolve to $2+2=1+3$, and commit the result} &
  (u = (+)\ 1\ 3 : \nat)(q = (=)\ t\ u : \prop) \\&
  (y = mark\ q\ \mathtt{``Also\ true}")
  \intertext{We could stop here: we have a nice object for commits,
    and a way to verify by typing that our repository is valid. Let's
    try to exploit the type system a little bit more: We  now
    introduce a higher-order object -- a
    patch maybe? -- taking a piece of syntax matching a certain form
    and constructing another object. In that case, it takes an
    equation and transforms it into an inequation:} &
  (P : (t\ u:\nat)(p=(=)\ t\ u:\prop)(q=(\leq)\ t\ u)\cdot mark\ q\
  \mathtt{``A\ fortiori}")
  \intertext{Now we can show that this patch applies to our
    repository by filling in the missing items:} &
  (z = P\ t\ u : (p=(=)\ t\ u:\prop)(q=(\leq)\ t\ u)\cdot mark\ q\
  \mathtt{``A\ fortiori}")
\end{align}
When the type of $P\ t\ u$ will be typed, the variable $p$ will be
given the same name as the previously-introduced $q$, because their
content match; whereas the new $q$ will be introduced into the
environment, and given a new unique name.
\\

The status of commits, patches etc. is still not very clear and the
previous example should be taken with a grain of salt, but this
hopefully gives a hint of what is possible with a dependently-typed,
memoizing language.
